The Real-time fMRI Neurofeedback Based Stratification of Default Network Regulation Neuroimaging Data Repository.

### 2016-10-08 mcdonald_nfb_datasharing_body.tex
Revised to reflect reviewer comments and added response letter

The present ongoing project is part of the Enhanced Nathan Kline Institute-Rockland Sample (NKI-RS) \cite{Nooner2012} which aims to capture deep and broad phenotyping of a large community-ascertained sample. In addition to the NKI-RS protocol, this project includes data collection from tasks that activate (Moral Dilemma task \cite{Harrison2008}) and deactivate the DMN (Multi-Source Interference Task \cite{Bush2006}), a resting state scan, a novel real-time fMRI neurofeedback-based paradigm that specifically probes DMN modulation \cite{Craddock2012}, and additional self-report measures.

@@

\subsection{Phenotypic data}
All phenotypic data are available as comma separated value (.csv) files from COINS. Summary scores calculated from the RVIP task info area available through COINs, trial-by-trial response information is available via NITRC (Data Citation Y1: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/RVIP-master.zip}).

\subsection{Imaging data}
Image headers have been wiped of protected health information and faces have been removed from the structural images. Data are available in NIfTI format arranged in the BIDS format \cite{Gorgolewski2016}. Acquisition parameters are provided in JSON files that are named the same as their corresponding imaging data. Task traces and responses, and physiological recordings, are also provided in JSON files along with the data. Additional details for all MRI data, as well as corresponding task information are available via NITRC (Data Citation X1: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/mri\_protocol.html\# scans-acquired}).

@@

\section{Contents of the Repository}
The Neurofeedback (NFB) repository contains neuroimaging and assessment data that characterizes DMN function in a community ascertained sample of adults (21-45 years old) with a variety of psychiatric diagnoses. The data is collected in a single-visit that occurs after the participants complete the Enhanced Nathan Kline Institute-Rockland Sample (NKI-RS) protocol \cite{Nooner2012}. The NKI-RS entails a 1 to 2 -visit deep phenotyping protocol \cite{Nooner2012} and a Connectomes oriented neuroimaging assessment (data collection visit schedules are available online at: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/sched.html}). NKI-RS phenotyping includes a variety of cognitive and behavioral assessments, a blood draw , a basic fitness assessment, nighttime actigraphy, medical history questionnaire, and the administration of the Structured Clinical Interview for DSM-IV-TR Non-patient edition \cite{First2002}. The NKI-RS imaging protocol includes structural MRI, diffusion MRI, several resting state fMRI, and perfusion fMRI scans. Although the NKI-RS data is being openly shared (\url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/}) it is not a part of the NFB repository and is not described in further detail herein.

@@

In addition to the NKI-RS protocol, participants completed a variety of assessments that probe cognitive, emotional, and behavioral domains that have been previously implicated with DMN function \cite{Andrews-Hanna2014,Hamilton2011,Sheline2009,Buckner2008}. These included the Affect Intensity Measure (AIM) \cite{Larsen1986}, Emotion Regulation Questionnaire (ERQ; \cite{Gross2003}), Penn State Worry Questionnaire (PSWQ; \cite{Meyer1990}), Perseverative Thinking Questionnaire (PTQ; \cite{Ehring2011}), Positive and Negative Affect Schedule – short version (PANAS-S; \cite{Watson1988}), Ruminative Responses Scale (RRS; \cite{Treynor2003}), and the Short Imaginal Process Inventory (SIPI; \cite{Huba1981}). Assessments were completed using web-based forms implemented in COINs \cite{Scott2011}. Sample mean, standard deviation, and range of the above measures for the first 125 participants are provided in Table 3.

@@

Several stages of online processing are necessary to enable the denoising of fMRI data and decoding DMN activity levels in real-time \cite{Craddock2012}. These stages include calculating transforms required to convert the DMN template from MNI space to subject space, creating white matter (WM) and cerebrospinal fluid (CSF) masks for extracting nuisance signals, and training a support vector regression (SVR) model for extracting DMN activity. The MRI session was optimized to collect the data required for these various processing steps, and to perform the processing, while minimizing delays in the experiment.
After acquiring a localizer, the scanning protocol began with the acquisition of a T1 weighted anatomical image used for calculating transforms to MNI space and white matter and CSF masks. Once the image was acquired it was transferred to a DICOM server on the real-time analysis computer (RTAC), which triggered initialization of online processing. The processing script started AFNI in real-time mode, configured it for fMRI acquisition, and began structural image processing. Structural processing included reorienting the structural image to RPI voxel order, skull-stripping using AFNI’s 3dSkullStrip \cite{Cox1996}, resampling the image to isotropic 2-mm voxels (to reduce computational cost of subsequent operations), segmentation into grey matter (GM), white matter (WM), and cerebrospinal fluid (CSF) using FSL’s FAST \cite{Zhang2001}, and normalization into MNI space using FSL’s FLIRT \cite{Jenkinson2002,Jenkinson2001}. WM and CSF probability maps were binarized using a 90\% threshold. The CSF mask was constrained to the lateral ventricles using an ROI from the AAL atlas to avoid overlap with GM.

@@

Next, a 6-minute resting state scan (182 volumes) was collected and used as training data to create the support vector regression (SVR) model. This procedure involved motion correction followed by a nuisance regression procedure to orthogonalize the data to six head motion parameters, mean WM, mean CSF, and global signal \cite{Friston1996,Fox2005,Lund2006}. A SVR model of the DMN was trained using a modified dual regression procedure in which a spatial regression to the canonical map of the DMN was performed to extract a time course of DMN activity. The Z-transformed DMN time course was then used as labels (independent variable), with the preprocessed resting state data as features (dependent variables), for SVR training (C=1.0, $\epsilon = 0.01$) using AFNI’s 3dsvm tool \cite{LaConte2005}. The result was a DMN map tailored to the individual participant based on preexisting expectations about DMN anatomy and function. After SVR training was completed, the MSIT (198 volumes), Moral Dilemma (144 volumes), and Neurofeedback test (412 volumes) scans were run. The order of the task based functional scans was counterbalanced and stratified for age and sex across participants.

@@

\begin{itemize}
    \item Standardized DVARS (DVARS): the mean intensity change between every pair of consecutive fMRI volumes, standardized to make it comparable between scanning protocols, smaller values are better \cite{Nichols2012}.
    \item Outliers: the mean number of outlier voxels found in each volume using AFNI’s 3dToutcount command, smaller values are better \cite{Cox1996}.
    \item Global correlation (gcorr): the average correlation between every pair of voxel time series in the brain, sensitive to physiological noise, head motion, as well as imaging technical artifacts, such as, signal bleeding in multi-band acquisitions, values that are closer to zero are better \cite{Saad2013}.
\end{itemize}

@@

Anatomical processing began with skull stripping using the BEaST toolset \cite{Eskildsen2012} with a study-specific library and careful manual correction of the results. The masks and BEaST library generated through this effort are shared through the Preprocessed Connectomes Project NFB Skullstripped repository (\url{http://preprocessed-connectomes-project.org/NFB_skullstripped/}) \cite{Puccio2016}. Skullstripped images were resampled to RPI orientation and then a non-linear transform between images and a 2mm MNI brain-only template (FSL, \cite{Smith2004}) was calculated using ANTs \cite{Avants2008}. The skullstripped images were additionally segmented into WM, GM, and CSF using FSL’s FAST tool \cite{Zhang2001}. A WM mask was calculated by applying a 0.96 threshold to the resulting WM probability map and multiplying the result by a WM prior map that was transformed into individual space using the inverse of the linear transforms previously calculated during the ANTs procedure. A CSF mask was calculated by applying a 0.96 threshold to the resulting CSF probability map and multiplying the result by a ventricle map derived from the Harvard-Oxford atlas distributed with FSL \cite{Makris2006}. The thresholds were chosen and the priors were used to avoid overlap with grey matter.

@@

>>>>

The Default Network Regulation Neuroimaging Repository contains data from a suite of fMRI experiments aimed at better understanding individual variation in DMN activity and modulation. Although it is a separate project, it has been harmonized with, and is distributed alongside, the Enhanced Nathan Kline Institute-Rockland Sample (NKI-RS) \cite{Nooner2012}, which aims to capture deep and broad phenotyping of a large community-ascertained sample. In addition to the NKI-RS protocol, this project includes data collection from tasks that activate (Moral Dilemma task \cite{Harrison2008}) and deactivate the DMN (Multi-Source Interference Task \cite{Bush2006}), a resting state scan, a novel real-time fMRI neurofeedback-based paradigm that specifically probes DMN modulation \cite{Craddock2012}, and additional self-report measures.

@@

\subsection{Phenotypic data}
Basic phenotypic data, which includes age, sex and handedness are available in the \textt{participants.tsv} file at the root of the repository. Comprehensive phenotypic data are available as comma separated value (.csv) files from COINS after completing a minimal data usage agreement. Summary scores calculated from the RVIP task info area available through COINs and trial-by-trial response information is available via NITRC (Data Citation Y1: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/RVIP-master.zip}).

\subsection{Imaging data}
The imaging data is released in unprocessed form except that image headers have been wiped of protected health information and faces have been removed from the structural images. Data are available in NIfTI files arranged in the BIDS format \cite{Gorgolewski2016}. Acquisition parameters are provided in JSON files that are named the same as their corresponding imaging data. Task traces (including stimulus onsets and durations) and responses, and physiological recordings, are also provided in JSON files along with the data. Additional details for all MRI data, as well as corresponding task information are available via NITRC (Data Citation X1: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/mri\_protocol.html\# scans-acquired}).

@@

\section{Contents of the Repository}
The Neurofeedback (NFB) repository contains neuroimaging and assessment data that characterizes DMN function in a community ascertained sample of adults (21-45 years old) with a variety of psychiatric diagnoses. The data is collected in a separate 2.5-hour visit that occurred within \textbf{**six**} months of completing the Enhanced Nathan Kline Institute-Rockland Sample (NKI-RS) protocol \cite{Nooner2012}. The NKI-RS entails a 1 to 2 -visit deep phenotyping protocol \cite{Nooner2012} and a Connectomes oriented neuroimaging assessment (data collection visit schedules are available online at: \url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/sched.html}). NKI-RS phenotyping includes a variety of cognitive and behavioral assessments, a blood draw , a basic fitness assessment, nighttime actigraphy, medical history questionnaire, and the administration of the Structured Clinical Interview for DSM-IV-TR Non-patient edition \cite{First2002}. The NKI-RS imaging protocol includes structural MRI, diffusion MRI, several resting state fMRI, and perfusion fMRI scans. Although the NKI-RS data is being openly shared (\url{http://fcon\_1000.projects.nitrc.org/indi/enhanced/}) it is not a part of the NFB repository and is not described in further detail herein.

@@

In addition to the NKI-RS protocol, participants completed a variety of assessments that probe cognitive, emotional, and behavioral domains that have been previously implicated with DMN function \cite{Andrews-Hanna2014,Hamilton2011,Sheline2009,Buckner2008}. These included the Affect Intensity Measure (AIM) \cite{Larsen1986}--to measure the strength or weakness with which one experiences both positive and negative emotions, Emotion Regulation Questionnaire (ERQ; \cite{Gross2003})--to assess individual differences in the habitual use of two emotion regulation strategies: cognitive reappraisal and expressive suppression, Penn State Worry Questionnaire (PSWQ; \cite{Meyer1990})--to measure worry, Perseverative Thinking Questionnaire (PTQ; \cite{Ehring2011})--to measure the broad idea of repetitive negative thought, Positive and Negative Affect Schedule – short version (PANAS-S; \cite{Watson1988})--to measure degrees of positive or negative affect, Ruminative Responses Scale (RRS; \cite{Treynor2003}) --to assess rumination that is related to, but not confounded by depression, and the Short Imaginal Process Inventory (SIPI; \cite{Huba1981})--to measure aspects of daydreaming style and content, mental style, and general inner experience. Assessments were completed using web-based forms implemented in COINs \cite{Scott2011}. Sample mean, standard deviation, and range of the above measures for the first 125 participants are provided in Table 3.

@@

The real time fMRI neurofeedback experiment utilizes a classifier based approach for extracting DMN activity levels from fMRI data TR-by-TR, similar to \cite{Craddock2012}. The classifier is trained from resting state fMRI data using a time course of DMN activity extracted from the data using spatial regression to a publicly available template derived from a meta-analysis of task and resting state datasets \cite{Smith2009,fmrib_RSNS}. Several stages of online processing are necessary to perform this classifier training, as well as, denoising of fMRI data in real-time. These stages include calculating transforms required to convert the DMN template from MNI space to subject space, creating white matter (WM) and cerebrospinal fluid (CSF) masks for extracting nuisance signals, and training a support vector regression (SVR) model for extracting DMN activity. The MRI session was optimized to collect the data required for these various processing steps, and to perform the processing, while minimizing delays in the experiment.

After acquiring a localizer, the scanning protocol began with the acquisition of a T1 weighted anatomical image used for calculating transforms to MNI space and white matter and CSF masks. Once the image was acquired it was transferred to a DICOM server on the real-time analysis computer (RTAC), which triggered initialization of online processing. The processing script started AFNI in real-time mode, configured it for fMRI acquisition, and began structural image processing. Structural processing included reorienting the structural image to RPI voxel order, skull-stripping using AFNI’s 3dSkullStrip \cite{Cox1996}, resampling the image to isotropic 2-mm voxels (to reduce computational cost of subsequent operations), segmentation into grey matter (GM), white matter (WM), and cerebrospinal fluid (CSF) using FSL’s FAST \cite{Zhang2001}, and normalization into MNI space using FSL’s FLIRT \cite{Jenkinson2002,Jenkinson2001}. WM and CSF probability maps were binarized using a 90\% threshold. The CSF mask was constrained to the lateral ventricles using an ROI from the AAL atlas to avoid overlap with GM.

@@

Next, a 6-minute resting state scan (182 volumes) was collected and used as training data to create the support vector regression (SVR) model. This procedure involved motion correction followed by a nuisance regression procedure to orthogonalize the data to six head motion parameters, mean WM, mean CSF, and global signal \cite{Friston1996,Fox2005,Lund2006}. A SVR model of the DMN was trained using a modified dual regression procedure in which a spatial regression to the unthresholded DMN template was performed to extract a time course of DMN activity. The Z-transformed DMN time course was then used as labels (independent variable), with the preprocessed resting state data as features (dependent variables), for SVR training (C=1.0, $\epsilon = 0.01$) using AFNI’s 3dsvm tool \cite{LaConte2005}. The result was a DMN map tailored to the individual participant based on preexisting expectations about DMN anatomy and function. After SVR training was completed (generally took less than 2 minutes), the MSIT (198 volumes), Moral Dilemma (144 volumes), and Neurofeedback test (412 volumes) scans were run. The order of the task based functional scans was counterbalanced and stratified for age and sex across participants.

@@

\begin{itemize}
    \item Standardized DVARS (DVARS): the mean intensity change between every pair of consecutive fMRI volumes, standardized to make it comparable between scanning protocols, smaller values are better \cite{Nichols2012}.
    \item Outliers: the mean number of outlier voxels found in each volume using AFNI’s 3dToutcount command, smaller values are better \cite{Cox1996}.
    \item Global correlation (gcorr): the average correlation between every pair of voxel time series in the brain, sensitive to physiological noise, head motion, as well as imaging technical artifacts, such as, signal bleeding in multi-band acquisitions, values that are closer to zero are better \cite{Saad2013}.
    \item Mean root mean square displacement (mean RMSD): the average distance between consecutive fMRI volumes \cite{Jenkinson99FD}. This has shown to be a more accurate representation of head motion than mean frame-wise displacement (meanFD) proposed by Power et. al. \cite{power2012} \cite{yan2013}.
\end{itemize}

@@

Anatomical processing began with skull stripping using the BEaST toolset \cite{Eskildsen2012} with a study-specific library and careful manual correction of the results. The masks and BEaST library generated through this effort are shared through the Preprocessed Connectomes Project NFB Skullstripped repository (\url{http://preprocessed-connectomes-project.org/NFB_skullstripped/}) \cite{Puccio2016}. Skullstripped images were resampled to RPI orientation and then a non-linear transform between images and a 2mm MNI brain-only template (FSL, \cite{Smith2004}) was calculated using ANTs \cite{Avants2008}. The skullstripped images were additionally segmented into WM, GM, and CSF using FSL’s FAST tool \cite{Zhang2001}. A WM mask was calculated by applying a 0.96 threshold to the resulting WM probability map and multiplying the result by a WM prior map (avg152T1\_white\_bin.nii.gz - distributed with FSL) that was transformed into individual space using the inverse of the linear transforms previously calculated during the ANTs procedure. A CSF mask was calculated by applying a 0.96 threshold to the resulting CSF probability map and multiplying the result by a ventricle map derived from the Harvard-Oxford atlas distributed with FSL \cite{Makris2006}. The thresholds were chosen and the priors were used to avoid overlap with grey matter.

@@

Group level analysis of the neurofeedback data indicates that the participants are able to modulate their DMN along with the task instructions. For all participants the group average time-course co-varies significantly with the task model and increases when removing participants who reported sleep. Greater than half of the  participants who reported no sleep had a  significant correlation (p \textless 0.05; r\textgreater 0.15, phase randomization permutation test) with the task model (see Fig. ~\ref{fig:nfb_perf}). The large portion of participants who were able to significantly modulate their DMN shows the effectiveness of the neurofeedback protocol.
